---
title: "Lecture 6: Good and Bad Confounds"
output: html_document
date: "2023-03-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(rethinking)
library(tidyverse)
theme_set(theme_bw())
```

## Take Aways of the week
### Fork: X ← Z → Y
- X and Y are associatied, but not directly each other.
- Z is common cause of X and Y 
- If we stratify on Z, X and Y are independent.
- Conditioning will block the path
- Example: `Marriage Rate` and `Divorce Rate` influenced by `Age of Marriage` 

### Pipe: X → Z → Y
- X and Y are associated. 
- Z transmitting causal relationship
- If we stratify Z, no association.
- Conditioning will block the path.
- Example: Fungus treatment and plant growth example. 
   
### Collider: X → Z ← Y. 
- X and Y are not associated
- They both influence Z
- Stratifying Z will make X and Y associated
- Conditioning on Z, opens the path, make it look like X and Y are correlated.
- Good looking and talented actors, stratifying on success  
  
### Descendent: 
Collider + a child of Z. Conditioning on the child will also act like collider.

### Backdoor criterion: 
1. List of all paths connecting X (var. interest) and Y (outcome)
2. Classify each path whether is open or closed. 
   - A path is open unless it contains a collider.
3. Check if there are backdoor paths (arrow entering X) 
4. If there are open backdoor paths, try to close it by conditioning.
  
## Question
The data in data(foxes) are 116 foxes from 30 different urban groups in England. 
```{r}
data(foxes)
head(foxes)

foxes %>% 
  ggplot(aes(y=1, x=weight, color=as.factor(group))) + 
  geom_point() +
  facet_wrap(.~group) +
  guides(color="none")
  
```

These fox groups
are like street gangs. Group size (groupsize) varies from 2 to 8 individuals.
Each group maintains its own (almost exclusive) urban territory. Some ter-
ritories are larger than others. The area variable encodes this information.
Some territories also have more avgfood than others. And food influences
the weight of each fox. Assume this DAG:

```{r, echo=FALSE, message=FALSE}
library(dagitty)
library(ggdag)
dagify(
  Food ~ Area,
  GrSize ~ Food,
  Weight ~ GrSize,
  Weight ~ Food
  ) %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point() + 
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag()
```

## 1. Use the backdoor criterion and estimate the total causal influence of A on
F. What effect would increasing the area of a territory have on the amount
of food inside it?

```{r}
library(rethinking)

d <- foxes

d$W <- standardize(d$weight)
d$A <- standardize(d$area)
d$F <- standardize(d$avgfood)
d$G <- standardize(d$groupsize)

model1 <- quap(
  alist(
  F ~ dnorm( mu , sigma ),
  mu <- a + bA*A,
  a ~ dnorm(0,10),
  bA ~ dnorm(0,10),
  sigma ~ dexp(1)
), data=d )

```
$$ Food_i = \alpha + \beta_A \times \text{Area}_i + \epsilon $$
```{r}
precis(model1)
coeftab_plot(coeftab(model1))
```

One SD of Area leads on average 0.88 SD in food availability.

## 2. Infer the total causal effect of adding food F to a territory on the weight
W of foxes. Can you calculate the causal effect by simulating an intervention
on food?
$$ Weight_i = \alpha + \beta_F \times \text{Food}_i + \epsilon $$

```{r}
model2 <- quap(
            alist(
              W ~ dnorm( mu , sigma ),
              mu <- a + bF*F,
              a ~ dnorm(0,1),
              bF ~ dnorm(0,1),
              sigma ~ dexp(1)
              ), data=d )

precis(model2)
coeftab_plot(coeftab(model2))
```

## 3. Infer the direct causal effect of adding food F to a territory on the weight
W of foxes. In light of your estimates from this problem and the previous
one, what do you think is going on with these foxes?
$$ Weight_i = \alpha + \beta_F \times \text{Food}_i + \beta_G \times \text{Group Size}_i + \epsilon $$
```{r}
model3 <- quap(
        alist(
          W ~ dnorm( mu , sigma ),
          mu <- a + bF*F + bG*G,
          a ~ dnorm(0,1),
          c(bF, bG) ~ dnorm(0,1),
          sigma ~ dexp(1)
          ), data=d )

precis(model3)
coeftab_plot(coeftab(model3))
```     

```{r}
model3b <- quap(
        alist(
          W ~ dnorm( mu , sigma ),
          mu <- a + bF*F + bG*G,
          a ~ dnorm(0,1),
          bF ~ dnorm(0,1),
          bG ~ dnorm(0,1),
          sigma ~ dexp(1)
          ), data=d )

coeftab_plot(coeftab(model3, model3b))
```

