---
title: "Lecture 5: Elemental Confounds"
output: html_document
date: "2023-03-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(rethinking)
library(tidyverse)
theme_set(theme_bw())
```
# Fork: X ← Z → Y
Z is common cause for both X and Y. 
```{r}
n <- 300
Z <- rbern(n, 0.5)
X <- rnorm(n, 2*Z -1)
Y <- rnorm(n, 2*Z -1)

```
Correlation between X and Y
```{r}
cor(X,Y)
```

Correlation when Z is 0
```{r}
cor(X[Z==0], Y[Z==0])
```

Correlation when Z is 1
```{r}
cor(X[Z==1], Y[Z==1])
```

```{r, message=FALSE, warning=FALSE}
df <- data.frame(X=X, Y=Y, Z=as.factor(Z)) 

df %>% 
ggplot(aes(x=X, y=Y, color=Z)) +
  geom_point(alpha=0.2) +
  geom_smooth(aes(x=X, y=Y), color="black", method="lm", se=FALSE) +
  geom_smooth(method="lm", se=FALSE) 
```
A ⫫/ 

# Marriage and Divorce Rates
Load data
```{r}
data("WaffleDivorce")

df <- WaffleDivorce %>% 
  mutate(D=standardize(Divorce),
         M=standardize(Marriage),
         A=standardize(MedianAgeMarriage))

df %>% 
  ggplot(aes(y=D, x=M)) +
  geom_point() +
  labs(x="Marriage Rate (standardized)", y="Divorce Rate(standardized)")

```


```{r dag2, out.width="50%"}
library(dagitty)
library(ggdag)
dagify(
  D ~ M,
  M ~ A,
  D ~ A
  ) %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point() + 
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag()
```

$$ \mu_i = \alpha + \beta_M M_i + \beta_A A_i $$

Often convenient to standardize data because: 
- Computation works better
- Easy to choose sensible priors


Generative model
$$ D_i \sim \mathcal{N}(\mu_i,\sigma)$$

$$ \mu_i = \alpha + \beta_m M_i + \beta_A A_i $$
$$ \alpha \sim \mathcal{N}(0,0.2)$$
$$ \beta_M \sim \mathcal{N}(0,0.5)$$
$$ \beta_A \sim \mathcal{N}(0,0.5)$$
$$ \sigma \sim \text{Exp}(1)$$

- Question in meeting: Why exponential distribution in sigma?



```{r}
model0 <- quap(
  alist(
    D ~ dnorm(mu,sigma),
    mu <- a + bM*M,
    a ~ dnorm(0,0.2),
    bM ~ dnorm(0,0.5),
    sigma ~ dexp(1)),
    data=df
  )

model1 <- quap(
  alist(
    D ~ dnorm(mu,sigma),
    mu <- a + bM*M + bA*A,
    a ~ dnorm(0,0.2),
    bM ~ dnorm(0,0.5),
    bA ~ dnorm(0,0.5),
    sigma ~ dexp(1)),
    data=df
  )

parameters <- precis(model1)
print(parameters)

coeftab_plot(coeftab(model0, model1))


```

alpha is centered on 0.


P(D|do(M)): distribution of D when we intervene on M when we do, deleting all arrows to M is deleted


```{r dag, out.width="50%"}
library(dagitty)
library(ggdag)
dagify(
  D ~ Mdo,
  #M ~ A,
  D ~ A
  ) %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point() + 
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag()
```

```{r}
post <- extract.samples(model1) # Extract samples from the posterior
n <- 1000 # sample 1000 observations (states)
As <- sample(df$A, size=n, replace = TRUE)

# simulate D for M=0 using the simulated Age
DM0 <- rnorm(n, post$a + post$bM * 0 + post$bA * As, post$sigma)

DM1 <- rnorm(n, post$a + post$bM * 1 + post$bA * As, post$sigma)

M10_contrast <- DM1 - DM0

data.frame(M10_contrast) %>% 
  ggplot(aes(x=M10_contrast)) +
  geom_density()
```

p(D|do(A)) 

# The Pipe X → Z → Y
- Z is a mediator.
- X and Y is associated. 
- Y not⫫ X
- Y ⫫ X | Z


```{r}
n <- 300
X <- rnorm(n)
Z <- rbern(n, inv_logit(X))
Y <- rnorm(n, 2*Z -1)

```
Correlation between X and Y
```{r}
cor(X,Y)
```

Correlation when Z is 0
```{r}
cor(X[Z==0], Y[Z==0])
```

Correlation when Z is 1
```{r}
cor(X[Z==1], Y[Z==1])
```
```{r, message=FALSE, warning=FALSE}
df <- data.frame(X=X, Y=Y, Z=as.factor(Z)) 

df %>% 
ggplot(aes(x=X, y=Y, color=Z)) +
  geom_point(alpha=0.2) +
  geom_smooth(aes(x=X, y=Y), color="black", method="lm", se=FALSE) +
  geom_smooth(method="lm", se=FALSE) 
```

